A Mechanical Formalization of Higher-Ranked Polymorphic Type Inference

### 2019-02-26 sections/Metatheory.tex
S4 except for 4.6 Abella

@@

The proof proceeds by induction on derivation of $\Gm\redto\nil$.
Proper applications of lemmas \ref{lem:insert} and \ref{lem:extract}
finish off the theorem.
\bruno{Give a bit more informal explanation here. Something like: If
  $\Gm$ is reducible then we can show that there is always an $\Om$
  where all the corresponding declarative judgements hold.}
\bruno{I wonder if 1 or 2 examples (concrete instantiations of the
  theorem) would be helpful to complement the explanation.}

@@

We prove completeness by induction on the derivation of $\Om\redto\nil$.
Since the declarative worklist is reduced judgment-by-judgment,
the induction always analyses the first judgment until it is satisfied.
As the algorithmic system introduces existential variables,
a declarative rule may correspond to multiple algorithmic rules,
and thus we analyse each of the possible cases.
\jimmy{Cite DK's proof as the sketch.}

@@

All other cases are handled easily... \jimmy{more details if necessary}

>>>>

The original DK's declarative system, shown in Figures \ref{fig:decl:sub} and \ref{fig:decl:typing},
has a few overlapping rules.
We propose a slightly modified declarative system, by changing a few rules in DK's declarative system:
\begin{enumerate}
\item Add a constraint to the subtyping rule $\mathtt{\forall L}$
$$
\inferrule*[right=$\mathtt{\forall L'}$]
{B \neq \all[b]B' \\ \Psi \vdash \tau \\ \Psi \vdash [\tau/a]A \le B}
{\Psi \vdash \all A \le B}
$$
\item Remove the redundant rule $\mathtt{Decl1I}$,
which could be easily derived by a combination of
$\mathtt{DeclSub}$, $\mathtt{Decl1I{\To}}$ and $\mathtt{{\le}Unit}$:
$$
\inferrule*[right=$\mathtt{DeclSub}$]
{
    \inferrule*[right=$\mathtt{Decl1I{\To}}$]
    {~}
    {\Psi \vdash () \To 1}
    \\
    \inferrule*[right=$\mathtt{{\le}Unit}$]
    {~}
    {\Psi \vdash 1 \le 1}
}
{\Psi \vdash () \Lto 1}
$$
\item Add constraints to the subsumption rule $\mathtt{DeclSub}$
$$
\inferrule*[right=$\mathtt{DeclSub'}$]
{e \neq \lam e' \\ A \neq \all A' \\ \Psi \vdash e \To A \\ \Psi \vdash A \le B}
{\Psi \vdash e \Lto B}
$$
\end{enumerate}

Now the new declarative system has no overlapping rules
and becomes more similar to the algorithmic system,
as the algorithmic rules contain similar constraints.
In addition, we proved that the modified declarative system is equivalent to the original one.

Modification (2) is relatively easy to justify, with the derivation given above:
the rule is redundant and can be replaced by a combination of three other rules.

Modificaitons (1) and (3) require inversion lemmas regarding rules that overlaps with them.
Rule $\mathtt{\forall L}$ initially overlaps with $\mathtt{\forall R}$ when facing judgment
$\Psi \vdash \all A \le \all[b]B$.
The following inversion lemma for Rule $\mathtt{\forall R}$ settles the overlapping:
\begin{lemma}[Invertibility of $\mathtt{\forall R}$]
If $\Psi \vdash A \le \all B$ then $\Psi, a \vdash A \le B$.
\end{lemma}
The lemma implies that giving a higher priority to $\mathtt{\forall R}$ does not affect the derivation.
Therefore the restriction $B \neq \all[b]B'$ is reasonable.

Rule $\mathtt{DeclSub}$ overlaps with both $\mathtt{Decl\forall I}$ and $\mathtt{Decl{\to}I}$.
Two similar inversion lemmas should be proven:
\begin{lemma}[Invertibility of $\mathtt{Decl\forall I}$]
If $\Psi \vdash e \Lto \all A$ then $\Psi, a \vdash e \Lto A$.
\end{lemma}
\begin{lemma}[Invertibility of $\mathtt{Decl{\to}I}$]
If $\Psi \vdash \lam e \Lto A \to B$ then $\Psi, x:A \vdash e \Lto B$.
\end{lemma}
These lemmas basically state that applying the specific rules is always no worse than
applying the more general Rule $\mathtt{DeclSub}$.
Proving both of the lemmas require an important property of the declarative system,
the subsumption lemma.
We first define the context subtyping relation in Figure~\ref{fig:context_subtyping},
and then show the subsumption lemma as follows.

\begin{figure}[t]
\framebox{$\Psi' \le \Psi$}
$$\begin{aligned}
\inferrule*[right=$\mathtt{CtxSubEmpty}$]
{~}{\nil \le \nil}\qquad
\inferrule*[right=$\mathtt{CtxSubTyVar}$]
{\Psi' \le \Psi}{\Psi', a \le \Psi, a}\qquad
\inferrule*[right=$\mathtt{CtxSubTmVar}$]
{\Psi' \le \Psi \\ \Psi \vdash A' \le A}{\Psi', x:A' \le \Psi, x:A}\qquad
\end{aligned}$$
\caption{Context Subtyping}
\label{fig:context_subtyping}
\end{figure}

\begin{lemma}[Subsumption]
Given $\Psi' \le \Psi$:
\begin{enumerate}
    \item If $\Psi \vdash e \Lto A$ and $\Psi \vdash A \le A'$ then $\Psi' \vdash e \Lto A$;
    \item If $\Psi \vdash e \To A$ then there exists
        $A'$ s.t. $\Psi \vdash A' \le A$ and $\Psi' \vdash e \To A$;
    \item If $\Psi \vdash \appInf{C}{e}{A}$ and $\Psi \vdash C' \le C$,
        then there exists $A'$ s.t. $\Psi \vdash A' \le A$ and $\Psi' \vdash \appInf{C'}{e}{A'}$.
\end{enumerate}
\end{lemma}

We tried to follow DK's proof on the subsumption lemma,
but the manual proof has several problems and is not able to be formalized.
Fortunately we fixed the issues and get the lemma proven.
Details about this issue can be found in Appendix~\ref{}.

\paragraph{Proof Sketch for Soundness and Completeness Theorems}
We now have three systems defined: the DK's declarative system,
the modified non-overlapping declarative system, and the algorithmic system.
First, we have got the statement that the non-overlapping declarative system,
as a restricted version, is equivalent to the DK's declarative system.
Then, the soundness of the algorithm can be directly shown against the original declarative system.
Finally, the completeness of the algorithm is shown against the non-overlapping declarative system,
which is equivalent the DK's declarative system.
As a result, we show that our algorithm is sound and complete with respect to DK's declarative system.

@@

The proof proceeds by induction on derivation of $\Gm\redto\nil$.
Interesting cases are the those involving existential variable instantiations,
including Rules 10, 11, 21 and 29.
Proper applications of lemmas \ref{lem:insert} and \ref{lem:extract}
help analyse the full instantiation of those existential variables.
For example, when $\al$ is solved to be $\al[1] \to \al[2]$ in the algorithm,
applying Extract gives two instantiations $\al[1] = \sigma$ and $\al[2] = \tau$.
Inserting back $\al = \sigma \to \tau$ for the induction case
matches the shape of induction hypothesis and thus finishes the corresponding case.

\bruno{Give a bit more informal explanation here. Something like: If
  $\Gm$ is reducible then we can show that there is always an $\Om$
  where all the corresponding declarative judgements hold.}
\bruno{I wonder if 1 or 2 examples (concrete instantiations of the
  theorem) would be helpful to complement the explanation.}

@@

We prove completeness by induction on the derivation of $\Om\redto\nil$.
Here we refer to the non-overlapping declarative system without chaning any syntax.
Since the declarative worklist is reduced judgment-by-judgment
(shown in Figure~\ref{fig:decl:worklist}),
the induction always analyses the first judgment until it is satisfied.
In Abella, we use a different relation to get the correct induction scheme.

As the algorithmic system introduces existential variables,
a declarative rule may correspond to multiple algorithmic rules,
and thus we analyse each of the possible cases.
\jimmy{Cite DK's proof as the sketch.}

@@

All other cases are relatively easy to prove.
The Insert and Extract lemmas are applied when the algorithm uses existential variables,
but transferred to a monotype for the declarative system,
such as Rules 6, 10, 11, 12-17, 21 and 29.
